---
title: "Epitope Logo Plot Comparison"
format: html
editor: visual
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ggseqlogo)
library(knitr)
```

## Load Data

```{r}
#| label: load-data
#| message: false

pos_viral_proteins <- read_csv("../data/pos_viral_proteins.csv")
pos_cancer         <- read_csv("../data/pos_EL_cancer_epitopes_hla0201.csv")
pos_healthy        <- read_csv("../data/pos_EL_healthy_epitopes_hla0201.csv")
```

## Data Descriptions

All three datasets share the following filters:

- **HLA restriction:** Only **HLA-A\*02:01** epitopes are included
- **Assay type:** **Eluted ligand data only**

Additional filters per dataset:

| Dataset | Source |
|---|---|
| **Viral** | Viral antigens only, human samples only, no mouse samples |
| **Cancer** | Epitopes associated with occurrence of cancer |
| **Healthy** | Epitopes from healthy individuals, no immunization |

---

# Peptide Length Distributions

```{r}
#| label: length-table-all

# Helper function to produce a length summary table
length_summary <- function(df, label) {
  df |>
    count(pep_length, name = "count") |>
    arrange(pep_length) |>
    mutate(
      percentage = round(count / sum(count) * 100, 2),
      dataset    = label
    )
}

bind_rows(
  length_summary(pos_viral_proteins, "Viral"),
  length_summary(pos_cancer,         "Cancer"),
  length_summary(pos_healthy,        "Healthy")
) |>
  select(dataset, pep_length, count, percentage) |>
  rename(
    "Dataset"        = dataset,
    "Peptide Length" = pep_length,
    "Count"          = count,
    "Percentage (%)" = percentage
  ) |>
  kable(caption = "Distribution of Peptide Lengths Across All Datasets")
```

---

# Logo Plots — Viral Epitopes

```{r}
#| label: filter-viral

target_lengths <- c(8, 9, 10, 11)

seqs_viral <- pos_viral_proteins |>
  filter(pep_length %in% target_lengths) |>
  split(f = ~pep_length) |>
  map(\(df) df$peptide)

map_dfr(seqs_viral, \(seqs) tibble(n = length(seqs)), .id = "Length") |>
  rename("Peptide Length" = Length, "Number of Sequences" = n) |>
  kable(caption = "Sequences per Logo Plot — Viral")
```

```{r}
#| label: logo-viral-8mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_viral[["8"]], method = "bits") +
  labs(
    title    = "Viral — 8-mers",
    subtitle = paste0("n = ", length(seqs_viral[["8"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-viral-9mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_viral[["9"]], method = "bits") +
  labs(
    title    = "Viral — 9-mers",
    subtitle = paste0("n = ", length(seqs_viral[["9"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-viral-10mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_viral[["10"]], method = "bits") +
  labs(
    title    = "Viral — 10-mers",
    subtitle = paste0("n = ", length(seqs_viral[["10"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-viral-11mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_viral[["11"]], method = "bits") +
  labs(
    title    = "Viral — 11-mers",
    subtitle = paste0("n = ", length(seqs_viral[["11"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

---

# Logo Plots — Cancer Epitopes

```{r}
#| label: filter-cancer

seqs_cancer <- pos_cancer |>
  filter(pep_length %in% target_lengths) |>
  split(f = ~pep_length) |>
  map(\(df) df$peptide)

map_dfr(seqs_cancer, \(seqs) tibble(n = length(seqs)), .id = "Length") |>
  rename("Peptide Length" = Length, "Number of Sequences" = n) |>
  kable(caption = "Sequences per Logo Plot — Cancer")
```

```{r}
#| label: logo-cancer-8mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_cancer[["8"]], method = "bits") +
  labs(
    title    = "Cancer — 8-mers",
    subtitle = paste0("n = ", length(seqs_cancer[["8"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-cancer-9mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_cancer[["9"]], method = "bits") +
  labs(
    title    = "Cancer — 9-mers",
    subtitle = paste0("n = ", length(seqs_cancer[["9"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-cancer-10mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_cancer[["10"]], method = "bits") +
  labs(
    title    = "Cancer — 10-mers",
    subtitle = paste0("n = ", length(seqs_cancer[["10"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-cancer-11mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_cancer[["11"]], method = "bits") +
  labs(
    title    = "Cancer — 11-mers",
    subtitle = paste0("n = ", length(seqs_cancer[["11"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

---

# Logo Plots — Healthy Epitopes

```{r}
#| label: filter-healthy

seqs_healthy <- pos_healthy |>
  filter(pep_length %in% target_lengths) |>
  split(f = ~pep_length) |>
  map(\(df) df$peptide)

map_dfr(seqs_healthy, \(seqs) tibble(n = length(seqs)), .id = "Length") |>
  rename("Peptide Length" = Length, "Number of Sequences" = n) |>
  kable(caption = "Sequences per Logo Plot — Healthy")
```

```{r}
#| label: logo-healthy-8mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_healthy[["8"]], method = "bits") +
  labs(
    title    = "Healthy — 8-mers",
    subtitle = paste0("n = ", length(seqs_healthy[["8"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-healthy-9mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_healthy[["9"]], method = "bits") +
  labs(
    title    = "Healthy — 9-mers",
    subtitle = paste0("n = ", length(seqs_healthy[["9"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-healthy-10mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_healthy[["10"]], method = "bits") +
  labs(
    title    = "Healthy — 10-mers",
    subtitle = paste0("n = ", length(seqs_healthy[["10"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

```{r}
#| label: logo-healthy-11mer
#| fig-width: 10
#| fig-height: 4
#| warning: false

ggseqlogo(seqs_healthy[["11"]], method = "bits") +
  labs(
    title    = "Healthy — 11-mers",
    subtitle = paste0("n = ", length(seqs_healthy[["11"]]), " peptides"),
    x = "Position", y = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

---

# Comparison — 9-mers Across All Datasets

```{r}
#| label: logo-comparison-9mer
#| fig-width: 10
#| fig-height: 8
#| warning: false

seqs_9mer_comparison <- list(
  "Viral"   = seqs_viral[["9"]],
  "Cancer"  = seqs_cancer[["9"]],
  "Healthy" = seqs_healthy[["9"]]
)

ggseqlogo(seqs_9mer_comparison, method = "bits", ncol = 1) +
  labs(
    title = "9-mer Sequence Logo Comparison",
    x     = "Position",
    y     = "Bits"
  ) +
  theme_logo() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    strip.text = element_text(size = 11, face = "bold")
  )
```

---

# Difference Logo Plots — 9-mer Comparisons

Difference logo plots show enrichment (positive, above line) and depletion 
(negative, below line) of amino acids at each position relative to a reference. 
Here each dataset is compared pairwise against the others.

```{r}
#| label: diff-logo-functions

# Helper: compute position probability matrix from sequences
get_ppm <- function(seqs) {
  # All amino acids
  aa <- c("A","C","D","E","F","G","H","I","K","L",
          "M","N","P","Q","R","S","T","V","W","Y")
  
  seq_matrix <- do.call(rbind, strsplit(seqs, ""))
  ppm <- apply(seq_matrix, 2, function(col) {
    counts <- table(factor(col, levels = aa))
    counts / sum(counts)
  })
  rownames(ppm) <- aa
  ppm
}

# Helper: compute difference PPM and plot
plot_diff_logo <- function(seqs_a, seqs_b, label_a, label_b) {
  ppm_a    <- get_ppm(seqs_a)
  ppm_b    <- get_ppm(seqs_b)
  diff_ppm <- ppm_a - ppm_b

  ggseqlogo(diff_ppm, method = "custom") +
    labs(
      title    = paste0("Difference Logo — ", label_a, " vs. ", label_b),
      subtitle = paste0(
        label_a, " (n = ", length(seqs_a), ") minus ",
        label_b, " (n = ", length(seqs_b), ")"
      ),
      x = "Position",
      y = "Difference in Probability"
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    theme_logo() +
    theme(
      plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 11)
    )
}
```

## Viral vs. Cancer

```{r}
#| label: diff-logo-viral-vs-cancer
#| fig-width: 10
#| fig-height: 4
#| warning: false

plot_diff_logo(
  seqs_viral[["9"]], seqs_cancer[["9"]],
  "Viral", "Cancer"
)
```

## Viral vs. Healthy

```{r}
#| label: diff-logo-viral-vs-healthy
#| fig-width: 10
#| fig-height: 4
#| warning: false

plot_diff_logo(
  seqs_viral[["9"]], seqs_healthy[["9"]],
  "Viral", "Healthy"
)
```

## Cancer vs. Healthy

```{r}
#| label: diff-logo-cancer-vs-healthy
#| fig-width: 10
#| fig-height: 4
#| warning: false

plot_diff_logo(
  seqs_cancer[["9"]], seqs_healthy[["9"]],
  "Cancer", "Healthy"
)
```